Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-30 paper2.qmd
section:all; process:write

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015]. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.

@@

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines Causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity [@Perron_et_al_2015], and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008]. This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. Two key observations support this conclusion: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- 3. But are our specific problems? -->
However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use its measurement scores for hypothesis testing. Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases). Moreover, because Thurstone's primary goal was to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269] and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], he did not address how to use the resulting measurement scores for hypothesis testing. Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978]. Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data. In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Bramley_2008; @Kelly_et_al_2022], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics. For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]. These misfit texts may exhibit larger discriminal dispersions or represent outliers--texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, the comparison of these texts results in more judgment discrepancies, yet the BTL model neither accounts for these cases nor offers any means to address them, except their elimination from the analysis.

@@

<!-- 5. So, what happens when the assumption does not hold? -->
Significant statistical and measurement issues can arise when the assumption of equal dispersions between stimuli does not hold. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
The correlation $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another. As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. A similar thought experiment, as the one depicted in @sec-theory-issue1a, can illustrate this idea. The experiment only assumes that the discriminal dispersions for both texts remain constant. @fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are positively correlated. As a result, the discriminal difference is less likely to favor Text B over Text A, as it is represented by the shaded gray area.

@@

<!-- 3. But does it hold?  -->
Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b]. @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.

@@

<!-- 3. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010]. For instance, when multiple texts are produced by the same individual, these texts are likely to share common features, such as writing style or even their general quality. Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient. For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.

@@

Finally, as discussed in the previous section, researchers risk discarding valuable information and introducing bias into trait estimates when they remove *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], considered outliers under the BTL model [@Grubbs_1969]. The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].

@@

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error). Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. The direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, demonstrating how neglecting measurement error $(\sigma_{T})$ by relying only on outcome averages can reduce the precision of a predictor’s estimated effect.

@@

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory). Additionally, the SCM incorporates a key design feature of CJ experiments: the judges' biases $(B_{kj})$. This extension builds on the arguments presented in @sec-theory-issue1b, contending that the discriminal difference becomes an observable outcome only through judges' perceptions. Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons.

Notably, all these variables and the outcome $(O_{iahbjk})$ constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$. Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

### The *conceptual population* data structure {#sec-theory-theoretical_P2}

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.

@@

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines Causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008]. This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the average relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the two hypothetical texts. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. Two key observations support this conclusion: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- 3. But are our specific problems? -->
However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement, and without providing clear guidance on how to use Case V's measurement scores for hypothesis testing. Specifically, he cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], as the case imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases). Moreover, because Thurstone's primary goal was to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269] and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], he did not address how to use the resulting measurement scores for hypothesis testing. Thus, given these limitations, it is surprising that CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978]. Furthermore, although the CJ practice has attempted to address the issue of hypothesis testing by using the scores' point estimates or their transformations, a critical question remains: Is this approach suitable for addressing these inquiries?

@@

<!-- 3. But does it hold in practice? -->
Notably, while assuming equal dispersions simplifies the trait measurement model, evidence from the CJ literature suggests that this assumption may fail to capture the complexity of modern CJ data. In particular, the assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a; @Bramley_2008; @Kelly_et_al_2022], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide indications of such discrepancies, particularly in the form of "misfit" statistics. For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004; @Pollitt_2012b; @Pollitt_2012a; @Goossens_et_al_2018]. These misfit texts may exhibit larger discriminal dispersions or represent outliers--texts with distinctive characteristics that deviate markedly from the rest of the sample [@Grubbs_1969]. In either case, the BTL model neither accounts for these cases nor offers any means to address them, except their elimination from the analysis.

@@

<!-- 5. So, what happens when the assumption does not hold? -->
Significant statistical and measurement issues can arise when the assumption of equal dispersions between stimuli does not hold. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Furthermore, if researchers acknowledge that misfit statistics may represent texts with different dispersions or outlying observations, the common CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020]. The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
The correlation $\rho$, illustrated in @fig-discriminal_difference, measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another. Similar to the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. A thought experiment, akin to the one presented in @sec-theory-issue1a, can illustrate this idea. Assuming that the discriminal dispersions for both texts remain constant, @fig-correlation shows that as the correlation between the two texts increases, the distribution of their discriminal difference becomes narrower. This narrowing, in turn, affects the probability that the discriminal difference distinctly favors Text B over Text A--denoted as $P(B > A)$--and thus directly influences the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated, as assumed in the BTL model $(\rho=0)$, the distribution of their discriminal difference is less narrow than in scenarios where the texts are positively correlated. As a result, it becomes less likely for the comparison to favor Text B over Text A, as indicated by the larger shaded area.

@@

<!-- 3. But does it hold?  -->
Despite these notable differences in the distribution of the discriminal difference under various correlational assumptions, in practice, experimental designs often adopt the assumption of no correlation between stimuli based on an old theoretical justification. Specifically, @Thurstone_1927b argued that stimuli could be treated as uncorrelated because judges' biases--arising from two opposing and equally weighted effects occurring during the pairwise comparisons--would cancel each other out. This idea was later formalized by @Andrich_1978, who provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, evidence from the CJ literature indicates that the assumption of zero correlation does not hold in practice in at least two scenarios: when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions or when additional hierarchical structures are relevant to the stimuli.

@@

<!-- 3. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli--a statistical phenomenon known as clustering [@Everitt_et_al_2010]. For instance, when the same individual produces multiple texts, those texts often exhibit common features, such as writing style or overall quality, that judges can easily recognize. In this regard, although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient. For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, similar to the ones discussed in @sec-theory-issue2.

@@

Finally, as discussed in the previous section, removing *misfit* judges--that is, judges whose assessments deviate markedly from the shared consensus [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], and considered outliers under the BTL model [@Grubbs_1969]--risk discarding valuable information and introducing bias into trait estimates. The direction and magnitude of these biases remain unpredictable, as they depend on which judges researchers exclude from the analysis [@Zimmerman_1994; @OHagan_2018; @McElreath_2020].

@@

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error). Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. The direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, showing that neglecting measurement error $(\sigma_{T})$--by relying solely on outcome averages--reduces the precision of a predictor's estimated effect $\beta_{X}$, as indicated by the solid lines.

@@

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ia}, T_{hb})$ and their discriminal difference $(D_{iahbjk})$ (see @sec-thurstone_theory). Additionally, the SCM incorporates a key design feature of CJ experiments: the judges' biases $(B_{kj})$. This extension builds on the arguments presented in @sec-theory-issue1b, contending that the discriminal difference becomes an observable outcome only through judges' perceptions. Given that such perceptions may be imperfect--and that each judge may carry some degree of bias [see @Pollitt_et_al_2003; @vanDaal_et_al_2016]--it is reasonable that judges' perceptions (bias) should be treated as an integral component of the CJ system from the outset, as this leads to a more accurate representation of the data-generating process underlying the pairwise comparisons. All these variables, together with the outcome $(O_{iahbjk})$, constitute the preliminary set of endogenous variables, $V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}$. Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

### The *conceptual-population* data structure {#sec-theory-theoretical_P2}
